\hspace{-0.1cm}{J-ReCoVer}:~Java~Reducer~Commutativity~Verifier

### 2019-06-24 draft.tex
Sections 4.1 and 4.2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Preprocessor and the Prover} \label{sec:preprocessor_prover}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Before entering the Prover component, the Preprocessor first performs a few program transformation tasks to simplify the verification task (Section~\ref{sec:program_trans1}). The output of the preprocessor is a commutativity-equivalent reducer program.  Then the algorithm of the Prover is explained in Section~\ref{sec:prover}.

%=============================================================================
\subsection{The Program Transformation Task in the Preprocessor}
\label{sec:program_trans1}
%=============================================================================

\subsubsection{Handling output inside the loop body} Recall that the reducer program can perform the ouput command $out(e)$ inside the loop body $s_2$. In this case, we remember the output value in a fresh variable $v$ and replace the $out(e)$ command with an assignment $v:=e$.
Notice that the value of $v$ stays the same from the point it has been assigned a value until the end of the loop body $s_2$.
If the values of $v$ are the same for all permutations of input lists at the end of $s_2$, we know that the reducer is commutative w.r.t. to all output values produced inside the loop body $s_2$.

\begin{wrapfigure}{r}{3.5cm}
	\vspace{-1.5cm}
	\begin{algorithm}[H]
		\;\;$m := \cur + 10$; \;
		\Loop{}{
			$t:=\cur$\;
			\uIf{ $t> m$}{
				$m := t$ \;
			}
		};
		$out(m)$\;\;
	\end{algorithm}
	\vspace{-0.8cm}
	\caption{Reducer max$^+$}
	\vspace{-0.8cm}
	\label{fig:reducer_max1}
\end{wrapfigure}

@@

\subsubsection{Handling the access to $\cur$ before entering the loop} In real-world examples, it is often that the $s_1$ part of the reducer also reads from the input. In the example in Figure~\ref{fig:reducer_max1}, the reducer program remembers the first input value in the variable $m$, increases its value by 10, and then updates its value to bigger ones if any occurs in the loop. The corresponding Equation~(\ref{eq:commu}) is valid in this case. However, the reducer is not commutative. A counterexample can be easily found. With the list $[1,2,3,4,5]$ and its permutation $[5,4,3,2,1]$ as the inputs, the reducer respectively outputs $11$ and $15$.

\begin{wrapfigure}{l}{5.3cm}
	\vspace{-1cm}
	\begin{algorithm}[H]
		$s:=1;$\;
		\Loop{}{
			\uIf{$s=1$}{$m := \cur + 10; s:= 2$}
			\uElse{
				$t:=\cur$\;
				\uIf{ $t > m$}{
					$m := t$ \;
				}
			}
		};
		$out(m)$
	\end{algorithm}	
	\vspace{-0.4cm}
	\caption{Reducer max$^{+\mathtt{fix}}$}
	\vspace{-0.4cm}
	\label{fig:reducer_max2}
\end{wrapfigure}
We handle the problem again by program transformation. To deal with the example in Figure~\ref{fig:reducer_max1}, we move the prefix $s_1$ into the loop body and use a new variable $s$ to force that the execution of $s_1$ is always before the original loop body. The result after the transformation is demonstrated in Figure~\ref{fig:reducer_max2}. The new reducer program over-approximates all possible inputs/outputs of the original one. When $s=1$, its output sequence is the same as the original reducer. When $s\neq 1$, it may produce outputs that are not possible by the original reducer.
It also satisfies all assumptions required for the Prover component:
(I) the commands before entering the loop does not have any $\cur$ expression,
(II) $out(v)$ command does not occur inside the loop body.
Since the new reducer has all behaviors of the original one, the new reducer being commutative implies that the original one is also commutative.

@@

This approach can be easily generalize to the situation where $\cur$ occurs multiple times in $s_1$. We first rewrite the $s_1$ part of the reducer $s_1;\rloop\{s_2\};s_3$ to the form
$$s'_0;x_{i_1}{:=} \cur;s'_1.\ldots x_{i_{m}}{:=}\cur;s'_m$$, where $\cur$ never occur in $s'_0,s'_1,\ldots, s'_m$. Then uses a branch statement with garud $s=j$ to handle the case $x_{i_j} := \cur; s'_j; s:=s+1$, for all $j \in [1,m]$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2mm}
\section{The Preprocessor and the Prover} \label{sec:preprocessor_prover}
\vspace*{-1mm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Before entering the Prover component, the Preprocessor first performs two
program transformations to simplify the verification task
(Section~\ref{sec:program_trans1}). The output of the preprocessor is a
commutativity-equivalent reducer program. The algorithm of the Prover is then
explained in Section~\ref{sec:prover}.

%=============================================================================
\vspace*{-2mm}
\subsection{The Program Transformations in the Preprocessor}
\label{sec:program_trans1}
\vspace*{-1mm}
%=============================================================================

%-----------------------------------------------------------------------------
\paragraph*{Handling the access to $\cur$ before entering the loop.}
%-----------------------------------------------------------------------------

\begin{wrapfigure}[11]{r}{3.5cm}
	\vspace{-17mm}
	\begin{algorithm}[H]
		\;\;$m := \cur + 10$; \;
		\Loop{}{
			$t:=\cur$\;
			\uIf{ $t> m$}{
				$m := t$ \;
			}
		};
		$out(m)$\;\;
	\end{algorithm}
	\vspace{-0.8cm}
	\caption{The max$^+$ reducer with input before the main loop.}
	\vspace{-12mm}
	\label{fig:reducer_max1}
\end{wrapfigure}

@@

In real-world reducers, it is often the case that the $s_1$ part of the reducer
reads from the input. Since our reduction of the commutativity problem to SMT
solving, which is presented in Section~\ref{sec:prover}, concentrates on the
influence of the input on the loop $s_2$ only, we need to transform the reducer
such that any input happens in the loop only.

In order to be able to explain the transformation that we perform on an example,
consider the example in Figure~\ref{fig:reducer_max1}. The reducer presented in
the figure remembers the first input value in the variable $m$, increases its
value by 10, and then updates its value to bigger ones if any occur in the loop.
The main loop of the reducer is commutative in this case, but the reducer is not
commutative. A counterexample can be easily found. With the list $[1,2,3,4,5]$
and its permutation $[5,4,3,2,1]$ as the inputs, the reducer outputs $11$ and
$15$, respectively.

\begin{wrapfigure}[12]{r}{5.3cm}
	\vspace{-9mm}
	\begin{algorithm}[H]
		$s:=1;$\;
		\Loop{}{
			\uIf{$s=1$}{$m := \cur + 10; s:= 2$}
			\uElse{
				$t:=\cur$\;
				\uIf{ $t > m$}{
					$m := t$ \;
				}
			}
		};
		$out(m)$
	\end{algorithm}	
	\vspace{-2mm}
	\caption{The max$^{+\mathtt{fix}}$ reducer.}
	\vspace{-4mm}
	\label{fig:reducer_max2}
\end{wrapfigure}

To deal with the problem, we use a transformation that we first explain on the
example from Figure~\ref{fig:reducer_max1}. In particular, we move the prefix
$s_1$ into the loop body and use a new variable $s$ to force that the execution
of $s_1$ is always before the original loop body. The result after the
transformation is demonstrated in Figure~\ref{fig:reducer_max2}. The new reducer
program over-approximates all possible inputs/outputs of the original one. When
$s=1$, its output sequence is the same as for the original reducer. When $s\neq
1$, it may produce outputs that are not possible by the original reducer.
However, since the new reducer has all behaviors of the original one, if the new
reducer is commutative, the original one is also commutative.

% TV: The below was removed since the requirements have not even be stated.
%
% It also satisfies all assumptions required for the Prover component: (1) the
% commands before entering the loop do not have any $\cur$ expression, (2) the
% $out(v)$ command does not occur inside the loop body.

@@

In general, the problem with the input before the loop can be handled as follows,
including the case where $\cur$ occurs multiple times in $s_1$. Assume that the
$s_1$ part of the reducer $s_1;\rloop\{s_2\};s_3$ has the form
$$s'_0;x_{i_1}{:=} \cur;s'_1; \ldots; x_{i_{m}}{:=}\cur;s'_m$$ where $\cur$ does
not occur in $s'_0,s'_1,\ldots, s'_m$. For all $j \in [1,m]$, we replace the
code $x_{i_j} := \cur; s'_j;$ from $s_1$ by a branch statement with the guard
$s=j$ and body $x_{i_j} := \cur; s'_j; s:=s+1$ in $s_2$. The original $s_2$ loop
body becomes the else branch of the newly introduced guard statement.

%-----------------------------------------------------------------------------
\subsubsection{Handling output inside the loop body} 
%-----------------------------------------------------------------------------

Another problem for our reduction to SMT is the fact that the reducer program
can perform the output command $out(e)$ inside the loop body $s_2$. Since our
SMT-based commutativity checking presented later on will be based on examining
possible values of the variables that appear in the $s_2$ loop body only (these
values then determine the output performed in $s_3$), we will replace each such
$out(e)$ command by the assignment $v:=e$ for a fresh variable $v$. This makes
the value that will be output visible for our analysis. The fact that we loose
the output command does not matter since---as already said---we will not track
the output commands but values of the variables that determine the value of the
output.  Notice that the value of $v$ stays the same from the point it has been
assigned a value until the end of the loop body $s_2$. If the value that $v$ can
have at the end of $s_2$ does not change for any permutation of the input lists,
we know that the reducer is commutative wrt all output values produced inside
the loop body $s_2$.
